Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2021-04-13 analysis/paper/paper.Rmd
TEXT EDITS AND ADD FORMAT DEPLOY TABLE INTEGERS

Haul-out behavior data from `r dplyr::n_distinct(deploy_tbl$speno)` bio-loggers deployed
on bearded, spotted, and ribbon seals was subset to include only records from 1 March to
15 July between `r min(lubridate::year(dat.sf$haulout_dt))` and
`r max(lubridate::year(dat.sf$haulout_dt))`. Bio-loggers were of the 'SPLASH' or 'SPOT'
family of tags developed by Wildlife Computers (Redmond, Washington, USA) and either
adhered to the hair on the seal or attached to the rear flipper inter-digital webbing. The
use of bio-loggers adhered to the back or head provide some benefits over flipper mounted
devices (e.g. increased satellite transmittal rates, locations at sea) but these are lost
during the annual molt, which limits the effective length of haul-out timelines.
Additionally, sensors attached to the head or dorsal region are often dry while the seal
is at the surface and can slightly bias the percent-dry values recorded by the device. For
this study, in cases where both tag types were deployed, percent dry observations from the
flipper tag were preferred. Sex and age class were determined at the time of
deployment by experts in the field. Seals determined to be less than one year were
classified as young-of-the-year. Subadults are those seals likely greater than one year of
age but not yet sexually mature. Adults are likely sexually mature and older than
approximately four years. For those tags deployed on young-of-the-year and transmitting
into the next year, the age class was advanced to subadult on 1 March of the following
year. Table \@ref(tab:tblDeploy) provides a summary of these deployments and data received
from them.

@@

deploy_tbl %>% 
  dplyr::group_by(species, sex, age) %>% 
  dplyr::summarise(
    n = n(),
    seal_hours = sum(n_hours)
  ) %>% 
  dplyr::mutate(
    deploy_stats = glue::glue("{n} ({seal_hours} seal hours)")
  ) %>% 
  dplyr::select(-c("n","seal_hours")) %>% 
  tidyr::pivot_wider(names_from = age, values_from = c(deploy_stats)) %>% 
  dplyr::ungroup() %>% 
  dplyr::rename(Species = species, Sex = sex, `Young of Year` = `YOUNG OF YEAR`,
                `Subadult` = SUBADULT, Adult = ADULT) %>% 
  flextable() %>% 
  add_header(`Young of Year` = "Age Class",
   `subadult` = "Age Class", Adult = "Age Class",
   top = TRUE ) %>% 
  merge_h(part = "header") %>% 
  align(align = "center", part = "header") %>% 
  font(fontname = "Source Serif Pro Light", part = "all") %>% 
  fontsize(size = 10) %>% 
  autofit()
```

@@

In addition to sex and age class (non-dependent *young-of-the-year*, sexually immature
*subadults*, and mature *adults*), we gathered additional variables that might help
explain variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and solar hour (for diurnal effects). Solar hour was calculated using the {solaR}
package [@perpinan2012jss] within the R statistical environment [@rcoreteam2019] using the
daily locations. We also compiled several meteorological variables shown to affect
haul-out behavior in other Arctic pinnipeds [@reder2003pb; @udevitz2009cjz; @perry2017]. In
particular, we linked locations from satellite tags to key weather values from the North
American Regional Reanalysis (NARR) model produced by the National Centers for
Environmental Prediction [@mesinger2006bams]. The NARR model assimilates observational
data to produce a long-term picture of weather over North America. Numerous weather
variables are made available across the region 8 times daily. For this study, NARR weather
values were subset to the extent of our study area over the Bering and Chukchi Seas at
3-hr intervals based on the native grid resolution of 32 km (1024 sq. km). The following
meteorological variables were interpolated and assigned to seal locations using a bilinear
method: 1) air temperature at 2m above the earth's surface, 2) wind consisting of
northerly and easterly vector components converted to wind speed using the Euclidean norm,
3) barometric pressure at sea level, and 4) precipitation (Table
\@ref(tab:tblCovariates)).

@@

```{r tblCovariates, include=TRUE, tab.id = "tblCovariates", tab.cap = "Explanatory covariates used in analyses of binary haul-out records for ice-associated seals."}
covariate_tbl <- tibble::tribble(
  ~Covariate, ~Type, ~Source, ~Description,
  "Age-sex class",   "Categorical", "Field Assessment", 
  "Classes included young-of-year, subadult, adult male and adult female. Young-of-year tags surviving past ~Feb 1 of the next year were treated as subadult in subsequent years.",
  "Hour", "Continuous; Fourier basis", "Bio-logger", 
  "We modeled solar hour using 6 variables of a Fourier-series basis to allow circularity, flexibility, and dimension reduction ",
  "Day",   "Continuous", "Bio-logger", 
  "Linear, quadratic, and cubic effects of day-of-year were modeled",
  "Precip", "Continuous", "NARR", 
  "Convective precipitation (kg/m2)",
  "Pressure", "Continuous", "NARR", 
  "Atmospheric pressure at sea level (kPa)",
  "Temp", "Continuous", "NARR",
  "Air temperatures predicted at 2m above the earth’s surface",
  "Wind", "Continuous", "NARR", 
  "Northerly and easterly vector components for wind were downloaded and converted into a single wind speed via the Euclidean norm.",
  "Northing", "Continuous", "Bio-logger","For bearded seals a measurement of northing was caclulated as latitude divided by the mean latitude across all locations"
)

@@

To properly account for temporal dependence within a computationally tractable framework,
we used generalized linear mixed pseudo-models (GLMPMs; [@verhoef2010cs]) to model
variation in haul-out behavior as a function of (1) covariate predictors, (2) temporally
autocorrelated random effects, and (3) individual random effects representing
heterogeneity in individual behavior. We used the {glmmLDTS} package [@verhoef2010cs] to
implement GLMPMs in the R programming environment . Owing to the large number of records,
we fit separate models to bearded, ribbon, and spotted seal data sets. We fit two
different models to our datasets. First, for each species, we fitted a year-independent
model that predicted average haul-out behavior as a function of demographic,
environmental, seasonal, and diurnal effects. Second, for ribbon and spotted seals (which
had considerably more data than bearded seals), we fitted models that included all the
effects from the first model, but also permitted annual variation in haul-out timing. This
second class of models was used to examine whether haul-out patterns varied by year and
to determine the annual timing of apparent peaks in haul-out behavior.

@@

A typical model fitting exercise would also include a model selection process. However,
AIC (and similar criteria) is not suitable when using pseudo-likelihoods. After fitting
GLMPM models, we used "type III" F-tests to calculate p-values [@verhoef2010cs] to
evaluate model performance and important terms. We also produced predictions of haul-out
behavior as a function of influential predictors (e.g. solar hour, day-of-year, age-sex).
Weather covariates for these predictions were based on daily or hourly smoothed weather
covariate values across the study region. Such predictions were then used to develop
haul-out probability surfaces, explore marginal effects of weather covariates, and
determine annual peaks in haul-out activity. The timing of peak haul-out behavior was
further used to regress against the annual maximum sea ice extent in the study region.

@@

To calculate plausible values for $\omega_{s,a}$, we obtained natural mortality predictions
from a hierarchical meta-analysis of phocid natural mortality [@trukhanova2018ee]. We then
assumed a 50/50 sex ratio at birth, 1 pup per litter, and reproductive schedules as
reported by Fedoseev [-@fedoseev2000] for seals in the western Bering Sea to formulate
Leslie matrices Ai for each species i. Stable age distributions
$v_i={v\_i0,v\_i1,⋯,v\_iA}$ were then determined by normalizing the dominant eigenvector
of $A_i$ so that they summed to one . Given a stable age distribution, we calculated the
proportion of the population by sex and stage (young-of-the-year: <1 year; subadult: at
least one year but sexually immature; adult: sexually mature) using data on sexual
maturity from Fedoseev [-@fedoseev2000]. In particular, the expected proportion of the
population that are sex $s$ and young-of-year was set as $\pi_{i,s,yoy}=0.5v_i0$, while
the expected proportion that are subadults and adults were set to

@@

  df <- data.frame(
    age_sex = a_s,
    solar_hour = rep(0:23, each = n_days),
    yday = rep(start_day:end_day, times = 24),
    northing = mean(data$northing),
    temp2 = temp_pred,
    wind = wind_pred,
    pressure = baro_pred,
    precip = precip_pred
  ) %>%
    mutate(
      sin1 = sin(pi * solar_hour / 12),
      cos1 = cos(pi * solar_hour / 12),
      sin2 = sin(pi * solar_hour / 6),
      cos2 = cos(pi * solar_hour / 6),
      sin3 = sin(pi * solar_hour / 4),
      cos3 = sin(pi * solar_hour / 4),
    ) %>%
    mutate(day = (yday - 120) / 10,
           day2 = day ^ 2,
           day3 = day ^ 3)
  df_list[[a_s]] <- df
  }
  if(length({{age_sex}}) > 1) {
  df_out <- bind_rows(df_list) %>%
    mutate(age_sex = forcats::fct_relevel(
      age_sex,c("ADULT.F","ADULT.M","SUBADULT","YOUNG OF YEAR"))
    )
  } else {
    df_out <- bind_rows(df_list)
  }
}
```

Models omitting year effects suggested that day-of-year, solar hour, age-sex class,
temperature, and wind substantially altered haul-out behavior of all three species, with F
tests producing p-values less than 0.05 for variables embodying these effects and/or their
interactions. Haul-out probabilities typically increased throughout March and April,
reaching a peak in May and early June before declining again. Diurnal patterns were
present, with maximum haul-out behavior centered around local solar noon.

>>>>

Haul-out behavior data from `r dplyr::n_distinct(deploy_tbl$speno)` bio-loggers deployed
on bearded, spotted, and ribbon seals was subset to include only records from 1 March to
15 July between `r min(lubridate::year(dat.sf$haulout_dt))` and
`r max(lubridate::year(dat.sf$haulout_dt))`. Bio-loggers were of the 'SPLASH' or 'SPOT'
family of tags developed by Wildlife Computers (Redmond, Washington, USA) and either
adhered to the hair on the seal or attached to the rear flipper inter-digital webbing. The
use of bio-loggers adhered to the back or head provide some benefits over flipper mounted
devices (e.g. increased satellite transmittal rates, locations at sea) but these are lost
during the annual molt, which limits the effective length of haul-out timelines.
Additionally, sensors attached to the head or dorsal region are often dry while the seal
is at the surface and can slightly bias the percent-dry values recorded by the device. For
this study, in cases where both tag types were deployed, percent dry observations from the
flipper tag were preferred. Sex and age class were determined at the time of deployment by
experts in the field. Field identification of age class can be uncertain in some
situations. The most notable is reliably discerning subadults from adults. In the case of
ribbon seals, pelage differences between subadults and adults are distinguishable, with
subadults often having less defined ribbons. Bearded seal subadults will often have a
spotted pattern in the pelage that is not seen in adults. Spotted seal pelage is
monomorphic and can not be used to discern age class. Despite these challenges, we feel
the age classifications used in this analysis are reliable. Seals determined to be less
than one year were classified as young-of-the-year. Subadults are those seals likely
greater than one year of age but not yet sexually mature. Adults are likely sexually
mature and older than approximately four years. For those tags deployed on
young-of-the-year and transmitting into the next year, the age class was advanced to
subadult on 1 March of the following year. Table \@ref(tab:tblDeploy) provides a summary
of these deployments and data received from them.

@@

deploy_tbl %>% 
  dplyr::group_by(species, sex, age) %>% 
  dplyr::summarise(
    n = n(),
    seal_hours = sum(n_hours)
  ) %>% 
  dplyr::mutate(
    deploy_stats = glue::glue('{n} ({format(seal_hours, big.mark=",")} seal hours)')
  ) %>% 
  dplyr::select(-c("n","seal_hours")) %>% 
  tidyr::pivot_wider(names_from = age, values_from = c(deploy_stats)) %>% 
  dplyr::ungroup() %>% 
  dplyr::rename(Species = species, Sex = sex, `Young-of-the-Year` = `YOUNG OF YEAR`,
                `Subadult` = SUBADULT, Adult = ADULT) %>% 
  flextable() %>% 
  add_header(`Young-of-the-Year` = "Age Class",
   Subadult = "Age Class", Adult = "Age Class",
   top = TRUE ) %>% 
  merge_h(part = "header") %>% 
  align(align = "center", part = "header") %>% 
  font(fontname = "Source Serif Pro Light", part = "all") %>% 
  fontsize(size = 10) %>% 
  autofit()
```

@@

In addition to sex and age class (non-dependent *young-of-the-year*, sexually immature
*subadults*, and mature *adults*), we gathered additional variables that might help
explain variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and local solar hour (for diurnal effects). Solar hour was calculated using the
{solaR} package [@perpinan2012jss] within the R statistical environment [@rcoreteam2019]
using the daily locations. We also compiled several meteorological variables shown to
affect haul-out behavior in other Arctic pinnipeds [@reder2003pb; @udevitz2009cjz;
@perry2017]. In particular, we linked locations from satellite tags to key weather values
from the North American Regional Reanalysis (NARR) model produced by the National Centers
for Environmental Prediction [@mesinger2006bams]. The NARR model assimilates observational
data to produce a long-term picture of weather over North America. Numerous weather
variables are made available across the region 8 times daily. For this study, NARR weather
values were subset to the extent of our study area over the Bering and Chukchi Seas at
3-hr intervals based on the native grid resolution of 32 km (1024 sq. km). The following
meteorological variables were interpolated and assigned to seal locations using a bilinear
method: 1) air temperature at 2m above the earth's surface, 2) wind consisting of
northerly and easterly vector components converted to wind speed using the Euclidean norm,
3) barometric pressure at sea level, and 4) precipitation (Table
\@ref(tab:tblCovariates)).

@@

```{r tblCovariates, include=TRUE, tab.id = "tblCovariates", tab.cap = "Explanatory covariates used in analyses of binary haul-out records for ice-associated seals."}
covariate_tbl <- tibble::tribble(
  ~Covariate, ~Type, ~Source, ~Description,
  "Age-sex class",   "Categorical", "Field Assessment", 
  "Classes included young-of-the-year, subadult, adult male and adult female. Young-of-the-year tags surviving past 1 March of the next year were treated as subadult in subsequent years.",
  "Hour", "Continuous; Fourier basis", "Bio-logger", 
  "Local solar hour using 6 variables of a Fourier-series basis to allow circularity, flexibility, and dimension reduction ",
  "Day",   "Continuous", "Bio-logger", 
  "Linear, quadratic, and cubic effects of day-of-year were modeled",
  "Precip", "Continuous", "NARR", 
  "Convective precipitation (kg/m2)",
  "Pressure", "Continuous", "NARR", 
  "Atmospheric pressure at sea level (kPa)",
  "Temp", "Continuous", "NARR",
  "Air temperatures predicted at 2m above the earth’s surface",
  "Wind", "Continuous", "NARR", 
  "Northerly and easterly vector components for wind were downloaded and converted into a single wind speed via the Euclidean norm.",
  "Northing", "Continuous", "Bio-logger","For bearded seals a measurement of northing was caclulated as latitude divided by the mean latitude across all locations"
)

@@

To properly account for temporal dependence within a computationally tractable framework,
we used generalized linear mixed pseudo-models (GLMPMs; [@verhoef2010cs]) to model
variation in haul-out behavior as a function of (1) covariate predictors, (2) temporally
autocorrelated random effects, and (3) individual random effects representing
heterogeneity in individual behavior. We used the {glmmLDTS} package [@verhoef2010cs] to
implement GLMPMs in the R programming environment [@rcoreteam2021]. Owing to the large
number of records, we fit separate models to bearded, ribbon, and spotted seal data sets.
We fit two different models to our datasets. First, for each species, we fitted a
year-independent model that predicted average haul-out behavior as a function of
demographic, environmental, seasonal, and diurnal effects. Second, for ribbon and spotted
seals (which had considerably more data than bearded seals), we fitted models that
included all the effects from the first model, but also permitted annual variation in
haul-out timing. This second class of models was used to examine whether haul-out patterns
varied by year and to determine the annual timing of apparent peaks in haul-out behavior.

@@

A typical model fitting exercise would also include a model selection process. However,
AIC (and similar criteria) is not suitable when using pseudo-likelihoods, because
pseudo-data generated in the model fitting process [@verhoef2010cs] differ between models
[@teneyck2017]. After fitting GLMPM models, we instead used "type III" F-tests to
calculate p-values [@verhoef2010cs] to evaluate model performance and important terms. We
also produced predictions of haul-out behavior as a function of influential predictors
(e.g. solar hour, day-of-year, age-sex). Weather covariates for these predictions were
based on daily or hourly smoothed weather covariate values across the study region. Such
predictions were then used to develop haul-out probability surfaces, explore marginal
effects of weather covariates, and determine annual peaks in haul-out activity. The timing
of peak haul-out behavior was further used to regress against the annual maximum sea ice
extent in the study region.

@@

To calculate plausible values for $\omega_{s,a}$, we obtained natural mortality
predictions from a hierarchical meta-analysis of phocid natural mortality
[@trukhanova2018ee]. We then assumed a 50/50 sex ratio at birth, 1 pup per litter, and
reproductive schedules as reported by Fedoseev [-@fedoseev2000] for seals in the western
Bering Sea to formulate Leslie matrices Ai for each species i. Stable age distributions
$v_i={v\_i0,v\_i1,⋯,v\_iA}$ were then determined by normalizing the dominant eigenvector
of $A_i$ so that they summed to one . Given a stable age distribution, we calculated the
proportion of the population by sex and stage (young-of-the-year: \<1 year; subadult: at
least one year but sexually immature; adult: sexually mature) using data on sexual
maturity from Fedoseev [-@fedoseev2000]. In particular, the expected proportion of the
population that are sex $s$ and young-of-year was set as $\pi_{i,s,yoy}=0.5v_i0$, while
the expected proportion that are subadults and adults were set to

@@

  df <- data.frame(
    age_sex = a_s,
    solar_hour = rep(0:23, each = n_days),
    yday = rep(start_day:end_day, times = 24),
    northing = mean(data$northing),
    temp2 = temp_pred,
    wind = wind_pred,
    pressure = baro_pred,
    precip = precip_pred
  ) %>%
    mutate(
      sin1 = sin(pi * solar_hour / 12),
      cos1 = cos(pi * solar_hour / 12),
      sin2 = sin(pi * solar_hour / 6),
      cos2 = cos(pi * solar_hour / 6),
      sin3 = sin(pi * solar_hour / 4),
      cos3 = sin(pi * solar_hour / 4),
    ) %>%
    mutate(day = (yday - 120) / 10,
           day2 = day ^ 2,
           day3 = day ^ 3)
  df_list[[a_s]] <- df
  }
  if(length({{age_sex}}) > 1) {
  df_out <- bind_rows(df_list) %>%
    mutate(age_sex = forcats::fct_relevel(
      age_sex,c("ADULT.F","ADULT.M","SUBADULT","YOUNG-OF-THE YEAR"))
    )
  } else {
    df_out <- bind_rows(df_list)
  }
}
```

Models omitting year effects suggested that day-of-year, solar hour, age-sex class,
temperature, and wind substantially altered haul-out behavior of all three species, with
$F$ tests producing $p$-values less than 0.05 for variables embodying these effects
and/or their interactions. Haul-out probabilities typically increased throughout March and
April, reaching a peak in May and early June before declining again. Diurnal patterns were
present, with maximum haul-out behavior centered around local solar noon.
